---
title: "Exploring the Shots Data from a Portion of the 2014-2015 NBA Season"
author: "Gitika Chaudhary, Jerry Hong, and Cashin Woo"
date: "12/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, lubridate, plotly, ggthemes)
```

## Information about the Data Set

This data set logs every shot during the NBA 2014-2015 regular season from October 2014-March 2015 (Important to note that this is not the span of the entire season) including a variety of factors that are relevant to the shot. We obtained this data from https://www.kaggle.com/dansbecker/nba-shot-logs, the data is scraped from NBA's REST API. 



```{r load data set}
shots <- read_csv("shot_logs.csv")
```

## A Brief Exploration of the Data Set

To get the gist of the data set, we will utilize the summary, structure, and head functions to depict the different aspects of the variables and observations the data set has.

```{r explore}
summary(shots)
str(shots)
head(shots)
```

After briefly exploring this data set, we notice there are 128069 observations with 21 variables. See the data dictionary below to see what each variable represents.Additionally, there is a mix of double and character data types throughout this data set. There is also a mix of numerical and categorical variables such as, measuring the shot distance as well as whether the game was at home or away. 

## Shot Log Data Dictionary

This dictionary describes what each variable of the data set represents.

GAME_ID

  Identification number of a particular NBA game.
  
MATCHUP

  The date the game occurred as well as the two teams that faced off.
  
LOCATION

  Whether the team was playing at home (H) or away (A).
  
W
  Whether the team won (W) or lost (L).
  
FINAL_MARGIN

  Final margin of victory or defeat for that game.
  
SHOT_NUMBER

  The numbered shot attempt by a player.
  
PERIOD

  The period in which the shot was attempted.
  
GAME_CLOCK

  The game clock at the time the shot was attempted.
  
SHOT_CLOCK

  The shot clock at the time the shot was attempted.
  
DRIBBLES

  The number of dribbles prior to the shot attempted.
  
TOUCH_TIME

  How long the player was holding the ball before shooting.
  
SHOT_DIST

  How far away, in feet, the player was when shooting the ball.
  
PTS_TYPE

  What kind of shot it was, either a 2 pointer or 3 pointer.
  
SHOT_RESULT

  Whether the shot missed or made.
  
CLOSEST_DEFENDER

  Who the closest defender was when the shot was taken.
  
CLOSEST_DEFENDER_PLAYER_ID

  Identification (number) assigned to the closest defender when the shot was taken.
  
CLOSE_DEF_DIST

  Distance of the closest defender to the person who took that shot (in feet).
  
FGM

  Number of shots scored by the shot.
  
PTS

  Number of points awarded for the shot.
  
Player_name

  Name of the player who took the shot.
  
Player_id	

  identification(number) assigned to the player who took the shot.


## Exploring the Different Shots Taken at Different Times

Our research questions we want to answer are what kind of shots are taken (in terms of distance from the basket) and how often they are being made at different times of the shot clock. In particular, we will analyze two players from the Golden State Warriors, Stephen Curry and Draymond Green. 


The first part of cleaning this data set is to separate the `MATCHUP` column into two separate columns. The original column has both the date and the two teams playing, so it makes sense to create a separate `DATE` column for that information. Moreover, we want to convert the `DATE` column into a date type column using `lubridate`.

```{r clean data set part I}
shots <- shots %>%
  separate(MATCHUP, into = c("DATE", "MATCHUP"), sep=" - ")

shots$DATE<-shots$DATE%>%mdy()

shots <- shots %>%
  mutate(TEAM = str_sub(shots$MATCHUP, 1, 3))

```


Another cleaning method we utilized is to reformat the `player_name` variable. The reason why we did this is because we want to make it consistent with the `CLOSEST_DEFENDER` column, which is formatted as "Last name, First name". Because `player_name` is formatted as "first last" we will reverse the 2 parts of the name and add a comma between them. The `rowwise()` function allows it to collapse the vector for each row of the data frame.  

```{r clean data set part II}
shots$player_name<-str_to_title(shots$player_name)
shots$player_name<-str_split(shots$player_name, pattern = " ")
shots$player_name<-lapply(X = shots$player_name,FUN = rev)

shots <- shots %>% 
  rowwise() %>% 
  mutate(player_name = str_c(player_name, collapse = ", "))


```


## Visualizing the Shot Map Based on Shot Clock

In order to analyze the shots taken and the time left on the shot clock for Stephen Curry and Draymond Green, we will utilize `plotly` so the user can hover over the data point and see the exact times and distances (Shot Clock, Shot Distance) at which the shot was taken. We will filter the data set by firstly the Golden State Warriors, and then Curry and Green. Moreover, we will color by the player so it is easier to visualize who shot which shot. 


```{r interactive plot}
shots2<-shots %>% 
  filter(TEAM == "GSW") %>%
  filter(player_name %in% c("Curry, Stephen", "Green, Draymond"))

plot_ly(shots2, x = ~ SHOT_CLOCK , y = ~ SHOT_DIST)%>%
    layout(title = "Curry vs. Green Shot Selection (2014-15)") %>%
    add_markers(color = shots2$player_name,
               text = ~paste0('Player: ', player_name))

```

Based on the visualization there are some features that stood out. Firstly, the spread and number of shots taken by Stephen Curry with respect to the distance; it goes to show that there are almost no kinds of shots he isn't afraid to take. Comparing it to a defensive player like Draymond Green, we can see that Green had far fewer mid-range shots and shot mostly from 3 and in the paint. An interesting cluster of points at the bottom right of the plot representing points scored by Draymond Green very close to the basket might attest to the fact that Green anticipates turnovers better than Curry; as a result, he is readily available to receive outlet passes and shoot in the paint at the start of the possession. 

## Visualizing the Shot Frequency Based on Distance from the Basket

The plot below displays a histogram representing the frequency of shots taken based on distance from the basket. We will create a facet plot for six players (Stephen Curry, LeBron James, James Harden, Kyrie Irving, DeMarcus Cousins and Anthony Davis). This can better observe how these players take different shots at different times of the shot clock.

```{r facet histogram}
shots %>%
  filter(player_name %in% c("Curry, Stephen", "James, Lebron", "Harden, James", "Irving, Kyrie", "Davis, Anthony", "Cousins, Demarcus")) %>%
  ggplot() + 
  geom_histogram(aes(x = SHOT_DIST, fill = player_name)) +
  labs(title = "Shot Usage by Distance Among Top NBA Players (2014-15)", 
       x = "Shot Distance (Feet)", 
       y = "Shot Usage") + 
  facet_wrap(~ player_name) + 
  theme_clean() + 
  theme(legend.position = "none")
```

Based on the figure above, we see that guards have a tendency to shoot from three more often than forwards. This is not surprising as the guards have different skill sets and responsibilities compared to forwards. What is interesting is that all players do not shoot from mid-range as often, which suggests the change in the game in recent years. In addition, James Harden shot more threes than Stephen Curry, despite the fact that Curry is most synonymous with 3-point shots. 

## Exploring the 2-pointer vs. 3-pointer Shots

We want to compare the efficacy between the 2-point and 3-point shots. Our goal is to answer what is the percentage of 2 pointers vs 3 pointers in the 2014-2015 season? Additionally, what type of shot is the most successful? 

### 2-pointers vs. 3-pointers Attempted in the 2014-15 season

In order to answer the first question, we first created another column `prop` which calculates the proportion of shots attempted between 2 and 3 pointers. Additionally, we factored the `PTS_TYPE` in order to make it discrete instead of continuous. Afterwards, we created a pie chart representing the proportions. 

```{r prop_shots}
shotcount <- shots %>%
  group_by(PTS_TYPE) %>%
  summarize(countmm = n()) %>%
  mutate(prop = countmm/sum(countmm))
shotcount

shotcount$PTS_TYPE<-factor(shotcount$PTS_TYPE)

ggplot(shotcount, aes(x="", y=prop, fill=PTS_TYPE)) +
  geom_bar(stat="identity", width=1) +
  ggtitle("Percentage of 2 point and 3 point Shots Attempted (2014-15)") +
  coord_polar("y", start=0)+theme_void() +
  scale_fill_brewer(palette="Set1")  

```

Based on the figure, 73.5 % of shots were 2 pointers in the 2014-2015 NBA season. 26.5 % of shots were 3 pointers. To further explore this data, we decided to see what percentage of 2 pointers and 3 pointers were actually made during the season. 

### 2-pointers vs. 3-pointers Made in the 2014-15 season

Because we're only focusing on shots that were made, we filtered the data set as such. The process afterwards is similar to as the previous question. The only difference is we created a bar graph comparing the two shot types. 

```{r prop_of_shots_made}
shotresult <- shots %>%
  filter(SHOT_RESULT=="made") %>%
  group_by(  SHOT_RESULT, PTS_TYPE) %>%
  summarize(countmm = n()) %>%
  mutate(prop = countmm/sum(countmm))

shotresult1 <- shotresult %>%
mutate(prop1=prop*100)
shotresult1
shotresult1$PTS_TYPE<-factor(shotresult1$PTS_TYPE)


ggplot(shotresult1, aes(x=PTS_TYPE, y=prop1,  fill =PTS_TYPE)) + 
  geom_bar(stat="identity" )+ ggtitle("Percentage of 2 point and 3 point Shots Made (2014-15)") +
  xlab("Type of Points") + ylab("Percentage")+ theme_economist()+
  theme(legend.title=element_blank())
```

Based on the figure, we found that 79.4% of all made shots were 2-pointers while 20.6% were 3-pointers throughout the season. This shows that despite the rise in the 3-point shot, the reliance on the 2-point shot is still strong. 
